\chapter[Pretty Printing]{Pretty Printing\\{\Large by Michael Kifer}}


This package provides methods for pretty printing the information about an
object or about all objects in a given class. This information can be saved
in a file or printed on the screen. This package must be first loaded into
a module, for instance, 
%% 
\begin{verbatim}
   ?- [prettyprint>>pp].
\end{verbatim}
%% 
and then the functionality of this package will be
accessible using the {\tt @pp} reference.

To pretty print information about an object, the following calls can be
used.  The first argument is the user module whose object is to be pretty
printed. (Recall that the same object can have completely different sets of
properties in different user modules, so the pretty printing methods need to
know which set of properties to use.)
%%
\begin{itemize}
\item {\tt ?Class[pp\_class]@pp} --- pretty print all objects in class {\tt
  ?Class} in the current module. Send the result to standard output.
\item {\tt ?Class[pp\_class(?Module)]@pp} --- same, but the information is
  printed on objects in module {\tt ?Module}. 
\item  {\tt ?Class[pp\_class(?Module,?Outfile)]@pp} --- same, but
  put the result in {\tt ?Outfile}.
\item {\tt ?Obj[pp\_self]} --- pretty print the state of the object {\tt ?Obj}
  in the current module. Send the result to standard output.
\item {\tt ?Obj[pp\_self(?Module)]@pp} --- same, but use the object {\tt ?Obj}
  in module {\tt ?Module}.  
\item {\tt ?Obj[pp\_self(?Module,?Outfile)]@pp} --- same, but send the result 
  to file {\tt ?Outfile}. 
\item {\tt ?Class[pp\_isa]@pp} --- pretty print the part of the isa
  hierarchy beneath {\tt ?Class} in the current module. Send the result to
  standard output.
\item {\tt ?Class[pp\_isa(?Module)]@pp} --- same, but use the isa hierarchy in
  module {\tt ?Module}. 
\item {\tt ?Class[pp\_isa(?Module,?Outfile)]@pp} --- same, but send the result
  to file {\tt ?Outfile}.
\end{itemize}
%%
The following example illustrates the use of this library:
%%
\begin{quote}
 {\tt
  flora2 ?- John[pp\_self(\thismodule)]@pp.
   }
\end{quote}
%%
When this method is called, the token \thismodule is replaced with the name
of the module in which the call occurs, so it known that it has to pretty
print the object {\tt John} in that module.



%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "ergo-packages"
%%% End: 
